Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-01-31 paper.tex
more reordering according to exposition
merge PDF

  \node[latent] at (1,3) (d){$\mathcal{D}$};
  \node[latent] at (2.5,3) (t){$\theta$};
  \node[latent] at (1,1.5) (z){$p$};
  \node[latent] at (3.5,1.5) (tx){$\theta_x$};
  \node[obs] at (1,-0.5) (x) {$x$};
  \edge {z}{x};
  \edge {d,t}{z};
  \plate {}{(tx)(z)(x)}{$x\in X$};
  \draw [->,dashed] (x.east) to[out = 0,in = -90] node(nn)[fill = white]{
          \includegraphics[width = 1.5cm]{figures/MLP.png} 
  } (tx.south);
  \node at (nn.south) {$q(x)$};
  \draw [->,dashed] (d.west) to[out = 180,in = 180] (z.west);
  \draw [->,dashed] (tx.west) -- (z.east);
  \end{tikzpicture}
  \caption{DSL $\mathcal{D}$ generates programs $p$ by sampling DSL primitives with probabilities $\theta$ (Algorithm~\ref{programGenerativeModel}). We observe program outputs $x$. A neural network $q(\cdot )$ called the \emph{recognition model} regresses from program outputs to a distribution over programs ($\theta_x = q(x)$). Solid arrows correspond to the top-down generative model. Dashed arrows correspond to the bottom-up recognition model.}\label{graphicalModel}
\end{figure}

@@

We maximize $J$ by alternatingly maximizing it w.r.t. the DSL and the frontiers:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.

@@

\textbf{Neural recognition model: tractably maxing $J$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
frontier $\mathcal{F}_x$ by searching for programs maximizing
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta_x$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This approach implements an amortized
inference scheme~\cite{ritchie2016deep} for the generative model in
Fig.~\ref{graphicalModel}.

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:% for using the simple strategy of enumerating programs:
\begin{itemize}
\item Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.
\item A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the
  search algorithm is very simple.  
\end{itemize}
A main drawback of an enumerative search algorithm is that we have no
efficient means of solving for arbitrary constants that might occur in the
program. In Section~\ref{regressionSection},
we will show how to find programs with real-valued constants
by automatically differentiating through the program and setting the constants using gradient descent.

\subsection{Learning a Neural Recognition Model}\label{recognitionSection}
The key idea here is that if the DSL is suitably fit to the domain then
we can get away with a simple, low-capacity model.

\begin{align}
  \mathcal{L}_{\text{RM}}& = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}\\
  \mathcal{L}_{\text{AE}}& = \expect_{x\sim X}\left[\sum_p Q_x(p)\log \probability [p|\mathcal{D},q(x)]\right]\nonumber\\
    \mathcal{L}_{\text{HM}}& = \expect_{p\sim(\mathcal{D},\theta) }\left[\log \probability[p|\mathcal{D},q(x)]\right],\text{ $p$ evaluates to $x$}\nonumber
\end{align}

@@

\begin{figure}
      \includegraphics[width = 9cm]{figures/polynomial.png}
\end{figure}
\subsection{String editing}
\begin{figure}

@@

Differentiate with respect to $\theta_e$ and set to zero 
\begin{align}
  \frac{c(x)}{\theta_x} &= N\frac{a(x)}{\sum_y a(y)\theta_y}
\end{align}
This equality holds if $\theta_x = c(x)/a(x)$:
\begin{align}
  \frac{c(x)}{\theta_x} &= a(x).\\
N\frac{a(x)}{\sum_y a(y)\theta_y}& = N\frac{a(x)}{\sum_y c(y)}   = N\frac{a(x)}{N} = a(x).
\end{align}
If this equality holds then $\theta_x \propto c(x)/a(x)$:
\begin{align}
  \theta_x = \frac{c(x)}{a(x)}\times \underbrace{\frac{\sum_y a(y)\theta_y}{N}}_{\text{Independent of $x$}}.
\end{align}

>>>>

  \node[latent] at (1,3) (d){$\mathcal{D}$};
  \node[latent] at (2.5,3) (t){$\theta$};
  \node[latent] at (1,1.5) (z){$p$};
  \node[latent] at (3.5,1.5) (tx){$\theta^{(x)}$};
  \node[obs] at (1,-0.5) (x) {$x$};
  \edge {z}{x};
  \edge {d,t}{z};
  \plate {}{(tx)(z)(x)}{$x\in X$};
  \draw [->,dashed] (x.east) to[out = 0,in = -90] node(nn)[fill = white]{
          \includegraphics[width = 1.5cm]{figures/MLP.png} 
  } (tx.south);
  \node at (nn.south) {$q(x)$};
  \draw [->,dashed] (d.west) to[out = 180,in = 180] (z.west);
  \draw [->,dashed] (tx.west) -- (z.east);
  \end{tikzpicture}
  \caption{DSL $\mathcal{D}$ generates programs $p$ by sampling DSL primitives with probabilities $\theta$ (Algorithm~\ref{programGenerativeModel}). We observe program outputs $x$. A neural network $q(\cdot )$ called the \emph{recognition model} regresses from program outputs to a distribution over programs ($\theta^{(x)} = q(x)$). Solid arrows correspond to the top-down generative model. Dashed arrows correspond to the bottom-up recognition model.}\label{graphicalModel}
\end{figure}

@@

We maximize $J$ by alternatingly maximizing it w.r.t. the DSL and the frontiers:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.

@@

\textbf{Neural recognition model: tractably maxing $J$ w.r.t. the
  frontiers.}  Here we train a neural network, $q$, to predict a
distribution over programs conditioned on a task. The objective of $q$
is to assign high probability to programs $p$ where
$\probability[x,p|\mathcal{D},\theta]$ is large.  With $q$ in hand we can find programs for
frontier $\mathcal{F}_x$ by searching for programs maximizing
$q(p|x)$.
The network $q$ exploits the structure of the DSL $\mathcal{D}$:
rather than directly predicting a distribution over $p$ conditioned on $x$,
it predicts a weight vector, $\theta^{(x)}$, and we define $q(p|x)\triangleq \probability[p|\mathcal{D},\theta = q(x)]$.
This approach implements an amortized
inference scheme~\cite{ritchie2016deep} for the generative model in
Fig.~\ref{graphicalModel}.

@@

Why enumerate, when the program synthesis community has invented many
sophisticated algorithms that search for programs?~\cite{solar2008program,schkufza2013stochastic,feser2015synthesizing,osera2015type,polozov2015flashmeta}.
We have two reasons:% for using the simple strategy of enumerating programs:
\begin{itemize}
\item Enumeration is a general approach that can be applied to any program induction problem. Many of these more sophisticated approaches require special conditions on
  the space of of programs.
\item A key point of our work is that learning the DSL, along with a neural recognition model, can make program induction tractable, even if the
  search algorithm is very simple.  
\end{itemize}
A main drawback of an enumerative search algorithm is that we have no
efficient means of solving for arbitrary constants that might occur in the
program. In Section~\ref{regressionSection},
we will show how to find programs with real-valued constants
by automatically differentiating through the program and setting the constants using gradient descent.
In Section~\ref{textSection}
we will show that the bottom-up neural recognition model can learn
which discrete constants should be included in a program.

\subsection{Learning a Neural Recognition Model}\label{recognitionSection}

The purpose of the recognition model is to accelerate the search over
programs.  It does this by learning to predict which programs both 
assign high likelihood to a task, and at the same time 
have high prior probability under the prior $\probability[\cdot |\mathcal{D},\theta]$.

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
can leverage the structure of the DSL,
and is \emph{not} responsible for
generating programs wholesale.
We will show that this lets us get away with a simple, low-capacity neural network.

We want a recognition model that closely approximates the true posteriors over programs, and so aim to minimize the following KL-divergence:
\begin{equation*}
  \expect\left[\text{KL}\left(\probability[p|x,\mathcal{D},\theta]||q(p|x) \right) \right]
\end{equation*}
which is equivalent to maximizing
$$
  \expect\left[\sum_p\probability[p|x,\mathcal{D},\theta]\log q(p|x) \right]
$$
  where the expectation is taken over tasks. One could take this expectation
  over the empirical distribution of the observations,
  like how an autoencoder is trained~\cite{hinton2006reducing}; or, one could take this expectation over samples from the generative model, like how a Helmholtz machine is trained~\cite{dayan1995helmholtz}.
  We found it useful to maximize both an autoencoder-style objective (written $\mathcal{L}_{\text{AE}}$) and a Helmholtz-style objective ($\mathcal{L}_{\text{HM}}$), giving the \system objective for a recognition model, $\mathcal{L}_{\text{RM}}$:
  \begin{align}
  \mathcal{L}_{\text{RM}}& = \mathcal{L}_\text{AE} + \mathcal{L}_\text{HM}\\
  \mathcal{L}_{\text{HM}}& = \expect_{p\sim(\mathcal{D},\theta) }\left[\log q(p|x)\right],\text{ $p$ evaluates to $x$}\nonumber\\
  \mathcal{L}_{\text{AE}}& = \expect_{x\sim X}\left[\sum_{p\in \mathcal{F}_x}
    \frac{\probability\left[x,p|\mathcal{D},\theta \right]}{\sum_{p'\in \mathcal{F}_x}\probability\left[x,p|\mathcal{D},\theta \right]}\log q(p|x)\right]\nonumber
  \end{align}

  Evaluating $\mathcal{L}_{\text{HM}}$ involves sampling programs from
  the current DSL, running them to get their outputs,
  and then training $q$ to regress from the outputs to the program.
  If these programs map inputs to outputs,
  then we need some way of sampling these inputs as well.
  Our solution to this problem is to sample the inputs
  from the empirical observed distribution of inputs in $X$.

@@

\begin{figure}
      \includegraphics[width = 9cm]{figures/polynomial.png}
\end{figure}
\subsection{String editing}\label{textSection}
\begin{figure}

@@

Differentiate with respect to $\theta_e$ and set to zero 
\begin{align}
  \frac{c(x)}{\theta^{(x)}} &= N\frac{a(x)}{\sum_y a(y)\theta_y}
\end{align}
This equality holds if $\theta^{(x)} = c(x)/a(x)$:
\begin{align}
  \frac{c(x)}{\theta_x} &= a(x).\\
N\frac{a(x)}{\sum_y a(y)\theta_y}& = N\frac{a(x)}{\sum_y c(y)}   = N\frac{a(x)}{N} = a(x).
\end{align}
If this equality holds then $\theta_x \propto c(x)/a(x)$:
\begin{align}
  \theta_x = \frac{c(x)}{a(x)}\times \underbrace{\frac{\sum_y a(y)\theta_y}{N}}_{\text{Independent of $x$}}.
\end{align}
